(Deep) Induction for GADTs

### 2021-04-05 induction-paper/GADT-induction.tex
Fix extra overline.
Comment on structural induction as consequence of deep induction.

@@

\subsection{(Deep) induction for $\mathsf{Equal}$}

@@

The induction rule for $\mathsf{Equal}$ is the type
\begin{multline*}
\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
%\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
\mathsf{\to CRefl\,P \to \forall (a : Set) (b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e}
\end{multline*}
We define a term $\mathsf{DIEqual}$ of such type as
\[
\mathsf{DIEqual\,P\,crefl\,a\,a\,Q_a\,Q_a'\,Refl\,L_E = crefl\,a\,Q_a\,Q_a'\,L_E}
\]
where
$\mathsf{crefl : CRefl\,P}$,
%$\mathsf{PRefl : \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}$
$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$
and $\mathsf{L_E : Equal^{\wedge}\,a\,a\,Q_a\,Q_a'\,Refl}$.
Having provided a well-defined term for it,
we have shown that the induction rule for $\mathsf{Equal}$ holds.

@@

Next, we shall provide an induction rule for the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
Again, the first step in deriving the induction rule for $\mathsf{Seq}$ consists in defining the predicate-lifting function over it, 
\[
\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
\]
which is given by pattern-matching as
\[
\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}
\]
where $\mathsf{Q_a : a \to Set}$ and $\mathsf{x : a}$,
and
\begin{multline*}
\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)} \\
\mathsf{=
\exists (Q_b : b \to Set) (Q_c : c \to Set)
\to Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e \times Seq^{\wedge}\,b\,Q_b\,s_b \times Seq^{\wedge}\,c\,Q_c\,s_c
}
\end{multline*}
where $\mathsf{e : Equal\,a\,(b \times c)}$,
$\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$.
We also need the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b : Set)(c : Set) \to b \times c}$, which is 
\[
\mathsf{Pair^{\wedge} : \forall (b : Set)(c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
\]
and it is defined as
\[
\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
\]
where $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$, $\mathsf{y : b}$ and $\mathsf{z : c}$.

Finally, let $\mathsf{CConst}$ be the type
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
\mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x)}
\end{multline*}
associated to the $\mathsf{Const}$ constructor,
and let $\mathsf{CSPair}$ be the type
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
\mathsf{\to \forall (a : Set) (b : Set) (c : Set)
	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c )}
\end{multline*}
associated to the $\mathsf{SPair}$ constructor,

@@

Consider such a generic GADT
\begin{equation}\label{eq:gadt}
\begin{array}{l}
\mathsf{data\, G\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (\ol{K\,\ol{b}})}
\end{array}
\end{equation}
which, for ease of notation,
we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{a}$)
and to have only one constructor $\mathsf{C}$.
Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty.
In the definition above, $\mathsf{F}$ has type $\mathsf{(Set \to Set) \to Set^{\beta} \to Set}$
and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
%The overline notation denotes a finite list:
%%and the vertical bars notation denotes the length of a list:
%%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
%each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
%and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
The constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (\ol{b : Set}) \to Equal\,a\,(K\,\ol{b}) \to F\,G\,\ol{b} \to G\,a}
\]
which is the form we shall use from now on.
%Notice that, in the above constructor, we have a type
%$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.

>>>>

{\color{blue} Moreover, for types admitting both deep induction and structural induction, the former generalizes the latter.
Indeed, structural induction rules can be derived from deep induction rules by choosing the constantly true predicate as the custom predicate traversing each data element contained in the data type.
This way, deep induction only inspects the structure of the data type and not its content, just like structural induction does.
A concrete example of this technique will be demonstrated in Section~\ref{sec:ind-equal}.}

@@

\subsection{(Deep) induction for $\mathsf{Equal}$}\label{sec:ind-equal}

@@

The induction rule for $\mathsf{Equal}$ is the type
\begin{multline*}
\mathsf{\forall (P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set)} \\
%\mathsf{\to \big( \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl \big)} \\
\mathsf{\to CRefl\,P \to \forall (a\,b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e: Equal\,a\,b)} \\
	\mathsf{\to Equal^{\wedge}\,a\,b\,Q_a\,Q_b\,e \to P\,a\,b\,Q_a\,Q_b\,e}
\end{multline*}
We define a term $\mathsf{DIEqual}$ of such type as
\[
\mathsf{DIEqual\,P\,crefl\,a\,a\,Q_a\,Q_a'\,Refl\,L_E = crefl\,a\,Q_a\,Q_a'\,L_E}
\]
where
$\mathsf{crefl : CRefl\,P}$,
%$\mathsf{PRefl : \forall (c : Set) (Q : c \to Set) (Q' : c \to Set) \to Equal^{\wedge}\,c\,c\,Q\,Q'\,Refl \to P\,c\,c\,Q\,Q'\,Refl}$
$\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$
and $\mathsf{L_E : Equal^{\wedge}\,a\,a\,Q_a\,Q_a'\,Refl}$.
Having provided a well-defined term for it,
we have shown that the induction rule for $\mathsf{Equal}$ holds.

The type $\mathsf{Equal}$ also has a standard structural induction rule $\mathsf{SIEqual}$,
\begin{multline*}
\mathsf{\forall (Q : \forall (a\,b : Set) \to Equal\,a\,b \to Set)} \\
\mathsf{\to \big( \forall (c : Set) \to P\,c\,c\,Refl \big)
	\to \forall (a\,b : Set) (e: Equal\,a\,b) \to P\,a\,b\,e}
\end{multline*}
As is the case for ADTs and nested types,
the structural induction rule for $\mathsf{Equal}$ is a consequence of the deep induction rule.
Indeed, we can define $\mathsf{SIEqual}$ as
\[
\mathsf{SIEqual\,Q\,srefl\,a\,b\,e = DIEqual\,P\,srefl\,a\,b\,K_1\,K_1\,e\,L_E}
\]
where $\mathsf{srefl: \forall (c : Set) \to P\,c\,c\,Refl}$
and $\mathsf{e : Equal\,a\,b}$, and
\begin{itemize}
\item $\mathsf{P : \forall (a\,b : Set) \to (a \to Set) \to (b \to Set) \to Equal\,a\,b \to Set}$ is defined as
$\mathsf{P\,a\,b\,Q_a\,Q_b\,e = Q\,a\,b\,e}$;
\item the left-most $\mathsf{K_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{a}$;
\item the right-most $\mathsf{K_1}$ is the constantly $\mathsf{1}$-valued predicate on $\mathsf{b}$;
\item $\mathsf{L_E : Equal^{\wedge}\,a\,b\,K_1\,K_1\,e}$ is defined by pattern matching,
i.e., in case $\mathsf{a = b}$ and $\mathsf{e}$ is $\mathsf{Refl}$,
it is defined as $\mathsf{L_E\,x = Refl : Equal\,a\,a}$.
\end{itemize}

That the structural induction rule is a consequence of the deep induction one is also true for all the examples below,
even though we will not remark it every time.

@@

Next, we shall provide an induction rule for the $\mathsf{Seq}$ type defined in Equation~\ref{eq:eq_seq}.
Again, the first step in deriving the induction rule for $\mathsf{Seq}$ consists in defining the predicate-lifting function over it, 
\[
\mathsf{Seq^{\wedge} : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set}
\]
which is given by pattern-matching as
\[
\mathsf{Seq^{\wedge}\,a\,Q_a\,(Const\,x) = Q_a\,x}
\]
where $\mathsf{Q_a : a \to Set}$ and $\mathsf{x : a}$,
and
\begin{multline*}
\mathsf{Seq^{\wedge}\,a\,Q_a\,(SPair\,b\,c\,e\,s_b\,s_c)} \\
\mathsf{=
\exists (Q_b : b \to Set) (Q_c : c \to Set)
\to Equal^{\wedge}\,a\, (b \times c)\, Q_a\, (Q_b \times Q_c) \, e \times Seq^{\wedge}\,b\,Q_b\,s_b \times Seq^{\wedge}\,c\,Q_c\,s_c
}
\end{multline*}
where $\mathsf{e : Equal\,a\,(b \times c)}$,
$\mathsf{s_b : Seq\,b}$ and $\mathsf{s_c : Seq\,c}$.
We also need the lifting of predicates over the polymorphic type of pairs, $\mathsf{Pair = \forall (b\,c : Set) \to b \times c}$, which is 
\[
\mathsf{Pair^{\wedge} : \forall (b\,c : Set) \to (b \to Set) \to (c \to Set) \to Pair\,b\,c \to Set}
\]
and it is defined as
\[
\mathsf{Pair^{\wedge}\,b\,c\,Q_b\,Q_c\,(y, z) = Q_b\,y \times Q_c\,z}
\]
where $\mathsf{Q_b : b \to Set}$, $\mathsf{Q_c : c \to Set}$, $\mathsf{y : b}$ and $\mathsf{z : c}$.

Finally, let $\mathsf{CConst}$ be the type
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
\mathsf{\to \forall (a : Set) (Q_a : a \to Set) (x : a) \to Q_a\,x \to P\,a\,Q_a\,(Const\,x)}
\end{multline*}
associated to the $\mathsf{Const}$ constructor,
and let $\mathsf{CSPair}$ be the type
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to Seq\,a \to Set)} \\
\mathsf{\to \forall (a\,b\,c : Set)
	(Q_a : a \to Set) (Q_b : b \to Set) (Q_c : c \to Set)} \\
	\mathsf{(s_b : Seq\,b) (s_c : Seq\,c) (e : Equal\,a\,(b \times c))
	\to Equal^{\wedge} a\, (b \times c)\, Q_a\, (Pair^{\wedge}\,b\,c\,Q_b\,Q_c)\, e} \\
	\mathsf{\to P\,b\,Q_b\,s_b \to P\,c\,Q_c\,s_c
	\to P a Q_a ( SPair\,b\,c\,e\,s_b\,s_c )}
\end{multline*}
associated to the $\mathsf{SPair}$ constructor,

@@

Consider such a generic GADT
\begin{equation}\label{eq:gadt}
\begin{array}{l}
\mathsf{data\, G\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C\, :\, F\,G\,\ol{b} \to G (K\,\ol{b})}
\end{array}
\end{equation}
which, for ease of notation,
we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{a}$)
and to have only one constructor $\mathsf{C}$.
Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty.
In the definition above, $\mathsf{F}$ has type $\mathsf{(Set \to Set) \to Set^{\beta} \to Set}$
and each $\mathsf{K}$ has type $\mathsf{Set^{\beta} \to Set}$.
%The overline notation denotes a finite list:
%%and the vertical bars notation denotes the length of a list:
%%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
%each of the $\mathsf{\alpha}$-many $\mathsf{K}$ depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
%and notice that the length of the list of types $\mathsf{\ol{a}}$ matches that of the list of type constructors $\mathsf{\ol{K}}$.
The constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (\ol{b : Set}) \to Equal\,a\,(K\,\ol{b}) \to F\,G\,\ol{b} \to G\,a}
\]
which is the form we shall use from now on.
%Notice that, in the above constructor, we have a type
%$\mathsf{Equal\,a\,(K\,\ol{b})}$ for each $\mathsf{a}$ and its corresponding type constructor $\mathsf{K}$.
